---
title: "STAT_380_Final_Project"
output: html_document
---

```{r}
#All Libraries
library(dplyr)
library(tidyverse)
library(cluster)
library(factoextra)
library(leaps)
library(glmnet)
library(ModelMetrics)
library(sjmisc)
```



```{r}
# For more infromation on the variables go to https://github.com/owid/covid-19-data/tree/master/public/data/vaccinations 
x <- read.csv("locations_vaccinations.csv")
y <- read.csv("vaccinations.csv")
z <- read.csv("us_state_vaccinations.csv")
CovidData <- readr::read_csv("owid-covid-data.csv.zip") 

# Note the CovidData is the data included in x, y, and z with other stuff. x y and z are there if you want something more specific without having to use the CovidData
```

```{r}
Vaccination <- read.csv("us_county_vaccinations.csv")
Population <- read.csv("us_county_populations.csv")
IncomeAndPoverty <- read.csv("us_county_income_and_poverty.csv")
Unemployment <- read.csv("us_county_unemployment.csv")
Geography <- read.csv("us_county_geography.csv")
WorldGeography <- read.csv("geography.csv")
```

Commented out portion below is for loading files and converting to csv files. The files would need to be downloaded for the code to work, so I commented it out to run the rest of the code more easily. Can also delete. 

<!-- ```{r} -->
<!-- # from https://data.cdc.gov/Vaccinations/COVID-19-Vaccinations-in-the-United-States-County/8xkx-amqh -->
<!-- COVID_19_Vaccinations_in_the_United_States_County <- file.choose() -->

<!-- Vaccination <- -->
<!--   data.table::fread(COVID_19_Vaccinations_in_the_United_States_County) -->

<!-- Vaccination <- -->
<!--   Vaccination %>% -->
<!--   rename(county = Recip_County, -->
<!--          state = Recip_State, -->
<!--          metro = Metro_status, -->
<!--          vaccinated = Series_Complete_Pop_Pct) %>% -->
<!--   filter(Date == "04/18/2022") %>%  # most recent data -->
<!--   select(county, state, metro, vaccinated) # variables of interest -->

<!-- write.csv(Vaccination, "us_county_vaccinations.csv") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # from https://www.census.gov/data/tables/time-series/demo/popest/2020s-counties-total.html -->
<!-- County_and_State_Annual_Resident_Population_Estimates_2021 <- file.choose() -->

<!-- Population <- -->
<!--   data.table::fread(County_and_State_Annual_Resident_Population_Estimates_2021) -->

<!-- # pattern to identify counties  -->
<!-- pattern <- "borough|census|city|county|district|municipality|parish" -->

<!-- # remove cases that are not counties -->
<!-- Population <-  -->
<!--   Population %>% -->
<!--   filter(grepl(pattern, CTYNAME, ignore.case = TRUE))  -->

<!-- Population <-  -->
<!--   Population %>% -->
<!--   rename(county = CTYNAME, -->
<!--          state = STNAME, -->
<!--          population = POPESTIMATE2020) %>% -->
<!--   select(county, state, population) %>% # variables of interest -->
<!--   mutate(state = state.abb[match(state, state.name)]) # convert names to abbreviations -->

<!-- write.csv(Population, "us_county_populations.csv") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # from https://www.census.gov/data/datasets/2020/demo/saipe/2020-state-and-county.html -->
<!-- US_State_and_County_Income_and_Poverty_Estimates_2020 <- file.choose() -->

<!-- IncomeAndPoverty <- -->
<!--   read_xls(US_State_and_County_Income_and_Poverty_Estimates_2020, range = "A4:Y3199") -->

<!-- # pattern to identify counties  -->
<!-- pattern <- "borough|census|city|county|district|municipality|parish" -->

<!-- # remove cases that are not counties -->
<!-- IncomeAndPoverty <-  -->
<!--   IncomeAndPoverty %>% -->
<!--   filter(grepl(pattern, Name, ignore.case = TRUE))  -->

<!-- IncomeAndPoverty <-  -->
<!--   IncomeAndPoverty %>% -->
<!--   rename(county = Name,  -->
<!--          state = `Postal Code`, -->
<!--          poverty = `Poverty Percent, All Ages`,  -->
<!--          income = `Median Household Income`) %>% -->
<!--   select(county, state, poverty, income) %>% # variables of interest -->
<!--   mutate(poverty = as.numeric(poverty), # convert plain text to number -->
<!--          income = as.numeric(income))  -->

<!-- write.csv(IncomeAndPoverty, "us_county_income_and_poverty.csv") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # from https://www.bls.gov/lau/ at "Labor force data by county, 2021 annual averages" under "County Data" -->
<!-- Labor_force_data_by_county_2021_annual_averages <- file.choose() -->

<!-- Unemployment <- -->
<!--   read_xlsx(Labor_force_data_by_county_2021_annual_averages, range = "A5:J3148") -->

<!-- Unemployment = Unemployment[-1, ] # delete empty first row -->

<!-- # pattern to extract state abbreviation -->
<!-- pattern <- ",\\s(.{2})" -->

<!-- Unemployment <-  -->
<!--   Unemployment %>% -->
<!--   # extract state abbreviation -->
<!--   tidyr::extract(`County Name/State Abbreviation`, into = "state", regex = pattern, remove = FALSE) %>%   -->
<!--   # remove state abbreviation  -->
<!--   mutate(`County Name/State Abbreviation` = str_remove(`County Name/State Abbreviation`, ",.*")) %>%  -->
<!--   rename(county = `County Name/State Abbreviation`, -->
<!--          unemployment = `(%)`) %>% -->
<!--   select(county, state, unemployment) # variables of interest -->

<!-- write.csv(Unemployment, "us_county_unemployment.csv") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # from https://public.opendatasoft.com/explore/dataset/us-county-boundaries/ -->
<!-- US_County_Boundaries <- file.choose() -->

<!-- Geography <- -->
<!--   data.table::fread(US_County_Boundaries) -->

<!-- Geography <- -->
<!--   Geography %>% -->
<!--   rename(county = NAMELSAD, -->
<!--          state = STUSAB, -->
<!--          latitude = INTPTLAT, -->
<!--          longitude = INTPTLON) %>% -->
<!--   select(county, state, latitude, longitude) -->

<!-- write.csv(Geography, "us_county_geography.csv") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # from https://github.com/gavinr/world-countries-centroids -->
<!-- World_Country_Centroids <- file.choose() -->

<!-- WorldGeography <- -->
<!--   data.table::fread(World_Country_Centroids) -->

<!-- WorldGeography <- -->
<!--   WorldGeography %>% -->
<!--   rename(country = COUNTRY) %>% -->
<!--   select(country, latitude, longitude) -->

<!-- write.csv(WorldGeography, "geography.csv") -->
<!-- ``` -->

```{r}
# compiling data sets for US county analysis

Vaccination <-
  Vaccination %>%
  # remove cases from Vaccination
  filter(!grepl("GU|PR|VI", state),
         !grepl("Unknown County", county))

Vaccination <-
  Vaccination %>%
  # remove and rename cases from Vaccination
  filter(!grepl("Valdez-Cordova Census Area", county)) %>%
  mutate(county = gsub("LaSalle", "La Salle", county))

Population <-
  Population %>%
  # remove and rename cases from Population
  filter(!grepl("Chugach Census Area|Copper River Census Area", county)) %>%
  mutate(county = gsub("Do\xf1a Ana County", "Dona Ana County", county),
         county = gsub("LaSalle", "La Salle", county),
         county = gsub("Petersburg Borough", "Petersburg Census Area", county))

Population[is.na(Population)] = "DC" # add state abbreviation for Washington, D.C.

IncomeAndPoverty <-
  IncomeAndPoverty %>%
  # remove and rename cases in IncomeAndPoverty
  filter(!grepl("Valdez-Cordova Census Area", county)) %>%
  mutate(county = gsub("Anchorage Borough", "Anchorage Municipality", county),
         county = gsub("DeBaca County", "De Baca County", county),
         county = gsub("De Kalb County", "DeKalb County", county),
         county = gsub("Juneau Borough", "Juneau City and Borough", county),
         county = gsub("Lagrange County", "LaGrange County", county),
         county = gsub("La Porte County", "LaPorte County", county),
         county = gsub("Mc Kean County", "McKean County", county),
         county = gsub("Petersburg Borough", "Petersburg Census Area", county),
         county = gsub("Sitka Borough", "Sitka City and Borough", county),
         county = gsub("Yakutat Borough", "Yakutat City and Borough", county))

Vaccination <-
  Vaccination %>%
  # remove case from Vaccination
  filter(!grepl("Kalawao County", county))

Unemployment <-
  Unemployment %>%
  # remove and rename cases in Unemployment
  filter(!grepl("Valdez-Cordova Census Area", county)) %>%
  mutate(county = gsub("Borough/municipality", "Municipality", county),
         county = gsub("County/city", "County", county),
         county = gsub("Borough/city", "City and Borough", county),
         county = gsub("LaSalle", "La Salle", county),
         county = gsub("County/town", "County", county),
         county = gsub("Petersburg Borough", "Petersburg Census Area", county))

Unemployment[is.na(Unemployment)] = "DC" # add state abbreviation for Washington, D.C.

USCovidData <-
  Vaccination %>%
  left_join(Population, by = c("county", "state")) %>%
  left_join(IncomeAndPoverty, by = c("county", "state")) %>%
  left_join(Unemployment, by = c("county", "state")) %>%
  unite("county_state", county:state, sep = " ") %>%
  select(county_state, vaccinated, poverty, population, income, unemployment) %>%
  rename(vaccination_rate = vaccinated) %>%
  na.omit() %>%
  distinct()
```

```{r}
# adding longitude and latitude to CovidData
CovidData <-
  CovidData %>%
  left_join(WorldGeography, by = c("location" = "country"))
```

```{r warning=FALSE}
Ratios <- CovidData %>%
  group_by(location) %>%
  summarise(vaccination_rate = max(total_vaccinations, na.rm = TRUE)/max(population, na.rm = TRUE), 
            gdp_per_capita = max(gdp_per_capita, na.rm = TRUE), 
            population_density = max(population_density, na.rm = TRUE), 
            human_development_index = max(human_development_index, na.rm = TRUE),
            latitude = max(latitude, na.rm = TRUE),
            longitude = max(longitude, na.rm = TRUE)) %>% 
  na.omit() %>%
  filter(is.finite(vaccination_rate),
         is.finite(gdp_per_capita),
         is.finite(population_density),
         is.finite(human_development_index),
         is.finite(latitude),
         is.finite(longitude))
Ratios
```

```{r}
# optimal number of clusters for countries
set.seed(10)
RatiosScaled <- Ratios %>% column_to_rownames(var = "location") %>% scale()
fviz_nbclust(RatiosScaled, kmeans, method = "wss")
```

```{r}
# PCA
res <- Ratios %>% column_to_rownames(var = "location") %>% prcomp(scale = TRUE)
get_eig(res)
df.res <- res$x
```

```{r}
# k-means clustering 
set.seed(10)
kc <- df.res %>% kmeans(7) 
summary(kc)
o <- order(kc$cluster)
data.frame(kc$cluster[o])
```

```{r warning=FALSE}
# PCA biplot
res %>% fviz_pca_biplot(label = "var",
                        habillage = as.factor(kc$cluster),
                        repel = "TRUE") +
  ylim(-2.5, 5)
```

PC1 places approximately equal weight on Human Development Index, Vaccination Rate. and Gross Domestic, Murder, and Rape, but less weight on UrbanPop. PC1 negatively correlates to rates of serious crimes. States with large negative scores on PC1, such as California, Nevada and Florida, have high crime rates, while states with positive scores on PC1, like North Dakota, have low crime rates. PC2 places most of its weight on UrbanPop and less weight on the other three variables. PC2 negatively correlates to the percentage of the population that lives in urban areas. States with large negative scores on PC2, like California, have a high percentage of their population living in urban areas, while states with positive scores on PC2, like Mississippi, have a low percentage of their population living in urban areas.

```{r}
# optimal number of clusters US counties
set.seed(1234)
USScaled <- USCovidData %>% column_to_rownames(var = "county_state") %>% scale()
fviz_nbclust(USScaled, kmeans, method = "wss")
```

```{r}
# PCA US counties
res2 <- USCovidData %>% column_to_rownames(var = "county_state") %>% prcomp(scale = TRUE)
get_eig(res2)
df.res2 <- res2$x
```

```{r}
# k-means clustering US counties
set.seed(1234)
kc2 <- df.res2 %>% kmeans(6) 
summary(kc2)
o <- order(kc2$cluster)
data.frame(kc2$cluster[o])
```

```{r warning=FALSE}
# PCA biplot US counties
res2 %>% fviz_pca_biplot(label = "var",
                         habillage = as.factor(kc2$cluster),
                         repel = "TRUE") +
  ylim(-10, 4)
```

```{r}
#Data Visualization
```

```{r}
#Hypothesis Testing
```

```{r warning=FALSE}
#Regression
df <- CovidData %>%
  group_by(location, continent) %>%
  summarise(vaccination_rate = (max(total_vaccinations, na.rm = TRUE)/max(population, na.rm = TRUE)) * 100,
            total_vaccinations = max(total_vaccinations, na.rm = TRUE),
            population = max(population, na.rm = TRUE),
            gdp_per_capita = max(gdp_per_capita, na.rm = TRUE), 
            life_expectancy = max(life_expectancy, na.rm = TRUE), 
            extreme_poverty = max(extreme_poverty, na.rm = TRUE), 
            human_development_index = max(human_development_index, na.rm = TRUE)) %>% 
  na.omit() %>%
  filter(is.finite(vaccination_rate),
         is.finite(total_vaccinations),
         is.finite(population),
         is.finite(gdp_per_capita),
         is.finite(life_expectancy),
         is.finite(extreme_poverty),
         is.finite(human_development_index))
head(df)  
df.cont1 = subset(df, select = -c(location, continent, total_vaccinations, population))
head(df.cont1)
```

```{r}
# forward selection
regfit.full.forw = regsubsets(vaccination_rate ~ ., data = df.cont1,  nvmax = 4, method="forward") 
summary(regfit.full.forw)
```

```{r}
reg.summary <- summary(regfit.full.forw) 
par(mfrow=c(2,2))
#adjr2 plot
plot(reg.summary$adjr2 ,xlab="Number of Variables ", ylab="Adjusted RSq",type="l")
max_adjr2 <- which.max(reg.summary$adjr2)
points(max_adjr2,reg.summary$adjr2[max_adjr2], col="red",cex=2,pch=20)
# AIC criterion (Cp) to minimize
plot(reg.summary$cp ,xlab="Number of Variables ",ylab="Cp", type='l')
min_cp <- which.min(reg.summary$cp )
points(min_cp, reg.summary$cp[min_cp],col="red",cex=2,pch=20)
# BIC criterion to minimize
plot(reg.summary$bic ,xlab="Number of Variables ",ylab="BIC",type='l')
min_bic <- which.min(reg.summary$bic)
points(min_bic,reg.summary$bic[min_bic],col="red",cex=2,pch=20)
```


```{r}
# backward selection
regfit.full.back = regsubsets(vaccination_rate ~ ., data = df.cont1,  nvmax = 4, method="backward") 
summary(regfit.full.forw)
```

```{r}
reg.summary <- summary(regfit.full.back) 
par(mfrow=c(2,2))
#adjr2 plot
plot(reg.summary$adjr2 ,xlab="Number of Variables ", ylab="Adjusted RSq",type="l")
max_adjr2 <- which.max(reg.summary$adjr2)
points(max_adjr2,reg.summary$adjr2[max_adjr2], col="red",cex=2,pch=20)
# AIC criterion (Cp) to minimize
plot(reg.summary$cp ,xlab="Number of Variables ",ylab="Cp", type='l')
min_cp <- which.min(reg.summary$cp )
points(min_cp, reg.summary$cp[min_cp],col="red",cex=2,pch=20)
# BIC criterion to minimize
plot(reg.summary$bic ,xlab="Number of Variables ",ylab="BIC",type='l')
min_bic <- which.min(reg.summary$bic)
points(min_bic,reg.summary$bic[min_bic],col="red",cex=2,pch=20)
```

```{r}
# stepwise selection
regfit.full.step = regsubsets(vaccination_rate ~ ., data = df.cont1,  nvmax = 4, method="seqrep") 
summary(regfit.full.step)
```

```{r}
reg.summary <- summary(regfit.full.step) 
par(mfrow=c(2,2))
#adjr2 plot
plot(reg.summary$adjr2 ,xlab="Number of Variables ", ylab="Adjusted RSq",type="l")
max_adjr2 <- which.max(reg.summary$adjr2)
points(max_adjr2,reg.summary$adjr2[max_adjr2], col="red",cex=2,pch=20)
# AIC criterion (Cp) to minimize
plot(reg.summary$cp ,xlab="Number of Variables ",ylab="Cp", type='l')
min_cp <- which.min(reg.summary$cp )
points(min_cp, reg.summary$cp[min_cp],col="red",cex=2,pch=20)
# BIC criterion to minimize
plot(reg.summary$bic ,xlab="Number of Variables ",ylab="BIC",type='l')
min_bic <- which.min(reg.summary$bic)
points(min_bic,reg.summary$bic[min_bic],col="red",cex=2,pch=20)
```

```{r} 
# summary from above steps # can be dropped
bestsub1=regsubsets(vaccination_rate~gdp_per_capita + life_expectancy + extreme_poverty + human_development_index,
                    data=df.cont1, nbest=16) 
plot(bestsub1, scale="adjr2") # the higher the better
plot(bestsub1, scale="Cp") # the lower the better
plot(bestsub1, scale="bic") # the lower the better
summary(bestsub1) # the top for each number set of variable is the best 
#Automatic selection - Forward
null=lm(vaccination_rate~1, data=df.cont1) #intercept only model; tilde 1
#full model- Regresses y on all variables in dataset
full=lm(vaccination_rate~., data=df.cont1)#full model- Regresses y on all variables in dataset
step(null, scope=list(lower=null, upper=full),
     direction="forward") # start from the intercept to the X1 to X5 
#Automatic selection - backward
step(full, data=modelsel, direction="backward") 
#Automatic selection - Stepwise 
step(null, scope = list(upper=full), data=df.cont1, direction="both") # forward but with re-evaluation
```

```{r}
attach(df.cont1)
lm.fit <- lm(vaccination_rate ~ life_expectancy + human_development_index , data = df.cont1)
summary(lm.fit) 
```

```{r}
# diagnostic tests
library(car)
vif(lm.fit)
par(mfrow=c(2,2))
plot(lm.fit)
```

```{r}
# take away 
# all three methods suggest the best model is one with predictors: life_expectancy + human_developement_index
# explain the covariance factors behind the results #
```