---
title: "STAT_380_Final_Project"
output: html_document
---

```{r}
#All Libraries
library(dplyr)
library(tidyverse)
library(cluster)
library(factoextra)
library(leaps)
library(glmnet)
library(ModelMetrics)
library(sjmisc)
```



```{r}
# For more infromation on the variables go to https://github.com/owid/covid-19-data/tree/master/public/data/vaccinations 
x <- read.csv("locations_vaccinations.csv")
y <- read.csv("vaccinations.csv")
z <- read.csv("us_state_vaccinations.csv")
CovidData <- readr::read_csv("owid-covid-data.csv.zip") 

# Note the CovidData is the data included in x, y, and z with other stuff. x y and z are there if you want something more specific without having to use the CovidData
```

```{r}
Vaccination <- read.csv("us_county_vaccinations.csv")
Population <- read.csv("us_county_populations.csv")
IncomeAndPoverty <- read.csv("us_county_income_and_poverty.csv")
Unemployment <- read.csv("us_county_unemployment.csv")
Geography <- read.csv("us_county_geography.csv")
WorldGeography <- read.csv("geography.csv")
```

Commented out portion below is for loading files and converting to csv files. The files would need to be downloaded for the code to work, so I commented it out to run the rest of the code more easily. Can also delete. 

<!-- ```{r} -->
<!-- # from https://data.cdc.gov/Vaccinations/COVID-19-Vaccinations-in-the-United-States-County/8xkx-amqh -->
<!-- COVID_19_Vaccinations_in_the_United_States_County <- file.choose() -->

<!-- Vaccination <- -->
<!--   data.table::fread(COVID_19_Vaccinations_in_the_United_States_County) -->

<!-- Vaccination <- -->
<!--   Vaccination %>% -->
<!--   rename(county = Recip_County, -->
<!--          state = Recip_State, -->
<!--          metro = Metro_status, -->
<!--          vaccinated = Series_Complete_Pop_Pct) %>% -->
<!--   filter(Date == "04/18/2022") %>%  # most recent data -->
<!--   select(county, state, metro, vaccinated) # variables of interest -->

<!-- write.csv(Vaccination, "us_county_vaccinations.csv") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # from https://www.census.gov/data/tables/time-series/demo/popest/2020s-counties-total.html -->
<!-- County_and_State_Annual_Resident_Population_Estimates_2021 <- file.choose() -->

<!-- Population <- -->
<!--   data.table::fread(County_and_State_Annual_Resident_Population_Estimates_2021) -->

<!-- # pattern to identify counties  -->
<!-- pattern <- "borough|census|city|county|district|municipality|parish" -->

<!-- # remove cases that are not counties -->
<!-- Population <-  -->
<!--   Population %>% -->
<!--   filter(grepl(pattern, CTYNAME, ignore.case = TRUE))  -->

<!-- Population <-  -->
<!--   Population %>% -->
<!--   rename(county = CTYNAME, -->
<!--          state = STNAME, -->
<!--          population = POPESTIMATE2020) %>% -->
<!--   select(county, state, population) %>% # variables of interest -->
<!--   mutate(state = state.abb[match(state, state.name)]) # convert names to abbreviations -->

<!-- write.csv(Population, "us_county_populations.csv") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # from https://www.census.gov/data/datasets/2020/demo/saipe/2020-state-and-county.html -->
<!-- US_State_and_County_Income_and_Poverty_Estimates_2020 <- file.choose() -->

<!-- IncomeAndPoverty <- -->
<!--   read_xls(US_State_and_County_Income_and_Poverty_Estimates_2020, range = "A4:Y3199") -->

<!-- # pattern to identify counties  -->
<!-- pattern <- "borough|census|city|county|district|municipality|parish" -->

<!-- # remove cases that are not counties -->
<!-- IncomeAndPoverty <-  -->
<!--   IncomeAndPoverty %>% -->
<!--   filter(grepl(pattern, Name, ignore.case = TRUE))  -->

<!-- IncomeAndPoverty <-  -->
<!--   IncomeAndPoverty %>% -->
<!--   rename(county = Name,  -->
<!--          state = `Postal Code`, -->
<!--          poverty = `Poverty Percent, All Ages`,  -->
<!--          income = `Median Household Income`) %>% -->
<!--   select(county, state, poverty, income) %>% # variables of interest -->
<!--   mutate(poverty = as.numeric(poverty), # convert plain text to number -->
<!--          income = as.numeric(income))  -->

<!-- write.csv(IncomeAndPoverty, "us_county_income_and_poverty.csv") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # from https://www.bls.gov/lau/ at "Labor force data by county, 2021 annual averages" under "County Data" -->
<!-- Labor_force_data_by_county_2021_annual_averages <- file.choose() -->

<!-- Unemployment <- -->
<!--   read_xlsx(Labor_force_data_by_county_2021_annual_averages, range = "A5:J3148") -->

<!-- Unemployment = Unemployment[-1, ] # delete empty first row -->

<!-- # pattern to extract state abbreviation -->
<!-- pattern <- ",\\s(.{2})" -->

<!-- Unemployment <-  -->
<!--   Unemployment %>% -->
<!--   # extract state abbreviation -->
<!--   tidyr::extract(`County Name/State Abbreviation`, into = "state", regex = pattern, remove = FALSE) %>%   -->
<!--   # remove state abbreviation  -->
<!--   mutate(`County Name/State Abbreviation` = str_remove(`County Name/State Abbreviation`, ",.*")) %>%  -->
<!--   rename(county = `County Name/State Abbreviation`, -->
<!--          unemployment = `(%)`) %>% -->
<!--   select(county, state, unemployment) # variables of interest -->

<!-- write.csv(Unemployment, "us_county_unemployment.csv") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # from https://public.opendatasoft.com/explore/dataset/us-county-boundaries/ -->
<!-- US_County_Boundaries <- file.choose() -->

<!-- Geography <- -->
<!--   data.table::fread(US_County_Boundaries) -->

<!-- Geography <- -->
<!--   Geography %>% -->
<!--   rename(county = NAMELSAD, -->
<!--          state = STUSAB, -->
<!--          latitude = INTPTLAT, -->
<!--          longitude = INTPTLON) %>% -->
<!--   select(county, state, latitude, longitude) -->

<!-- write.csv(Geography, "us_county_geography.csv") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # from https://github.com/gavinr/world-countries-centroids -->
<!-- World_Country_Centroids <- file.choose() -->

<!-- WorldGeography <- -->
<!--   data.table::fread(World_Country_Centroids) -->

<!-- WorldGeography <- -->
<!--   WorldGeography %>% -->
<!--   rename(country = COUNTRY) %>% -->
<!--   select(country, latitude, longitude) -->

<!-- write.csv(WorldGeography, "geography.csv") -->
<!-- ``` -->

```{r}
# compiling data sets for US county analysis

Vaccination <-
  Vaccination %>%
  # remove cases from Vaccination
  filter(!grepl("GU|PR|VI", state),
         !grepl("Unknown County", county))

Vaccination <-
  Vaccination %>%
  # remove and rename cases from Vaccination
  filter(!grepl("Valdez-Cordova Census Area", county)) %>%
  mutate(county = gsub("LaSalle", "La Salle", county))

Population <-
  Population %>%
  # remove and rename cases from Population
  filter(!grepl("Chugach Census Area|Copper River Census Area", county)) %>%
  mutate(county = gsub("Do\xf1a Ana County", "Dona Ana County", county),
         county = gsub("LaSalle", "La Salle", county),
         county = gsub("Petersburg Borough", "Petersburg Census Area", county))

Population[is.na(Population)] = "DC" # add state abbreviation for Washington, D.C.

IncomeAndPoverty <-
  IncomeAndPoverty %>%
  # remove and rename cases in IncomeAndPoverty
  filter(!grepl("Valdez-Cordova Census Area", county)) %>%
  mutate(county = gsub("Anchorage Borough", "Anchorage Municipality", county),
         county = gsub("DeBaca County", "De Baca County", county),
         county = gsub("De Kalb County", "DeKalb County", county),
         county = gsub("Juneau Borough", "Juneau City and Borough", county),
         county = gsub("Lagrange County", "LaGrange County", county),
         county = gsub("La Porte County", "LaPorte County", county),
         county = gsub("Mc Kean County", "McKean County", county),
         county = gsub("Petersburg Borough", "Petersburg Census Area", county),
         county = gsub("Sitka Borough", "Sitka City and Borough", county),
         county = gsub("Yakutat Borough", "Yakutat City and Borough", county))

Vaccination <-
  Vaccination %>%
  # remove case from Vaccination
  filter(!grepl("Kalawao County", county))

Unemployment <-
  Unemployment %>%
  # remove and rename cases in Unemployment
  filter(!grepl("Valdez-Cordova Census Area", county)) %>%
  mutate(county = gsub("Borough/municipality", "Municipality", county),
         county = gsub("County/city", "County", county),
         county = gsub("Borough/city", "City and Borough", county),
         county = gsub("LaSalle", "La Salle", county),
         county = gsub("County/town", "County", county),
         county = gsub("Petersburg Borough", "Petersburg Census Area", county))

Unemployment[is.na(Unemployment)] = "DC" # add state abbreviation for Washington, D.C.

USCovidData <-
  Vaccination %>%
  left_join(Population, by = c("county", "state")) %>%
  left_join(IncomeAndPoverty, by = c("county", "state")) %>%
  left_join(Unemployment, by = c("county", "state")) %>%
  unite("county_state", county:state, sep = " ") %>%
  select(county_state, vaccinated, poverty, population, income, unemployment) %>%
  na.omit() %>%
  distinct()
```

```{r}
# adding longitude and latitude to CovidData
CovidData <-
  CovidData %>%
  left_join(WorldGeography, by = c("location" = "country"))
```

```{r warning=FALSE}
Ratios <- CovidData %>%
  group_by(location) %>%
  summarise(ratio = max(total_vaccinations, na.rm = TRUE)/max(population, na.rm = TRUE), 
            gdp = max(gdp_per_capita, na.rm = TRUE), 
            indexS = max(stringency_index, na.rm = TRUE), 
            density = max(population_density, na.rm = TRUE), 
            indexHD = max(human_development_index, na.rm = TRUE),
            latitude = max(latitude, na.rm = TRUE),
            longitude = max(longitude, na.rm = TRUE)) %>% 
  na.omit() %>%
  filter(is.finite(ratio),
         is.finite(gdp),
         is.finite(indexS),
         is.finite(density),
         is.finite(indexHD),
         is.finite(latitude),
         is.finite(longitude))
Ratios
```

```{r}
# optimal number of clusters for countries
set.seed(1234)
RatiosScaled <- Ratios %>% column_to_rownames(var = "location") %>% scale()
fviz_nbclust(RatiosScaled, kmeans, method = "wss")
```

```{r}
# PCA
res <- Ratios %>% column_to_rownames(var = "location") %>% prcomp(scale = TRUE)
get_eig(res)
df.res <- res$x
```

```{r}
# k-means clustering 
set.seed(1234)
kc <- df.res %>% kmeans(6) 
summary(kc)
o <- order(kc$cluster)
data.frame(kc$cluster[o])
```

```{r warning=FALSE}
# PCA biplot
res %>% fviz_pca_biplot(label = "var",
                        habillage = as.factor(kc$cluster),
                        repel = "TRUE") +
  ylim(-5, 3)
```

```{r}
# optimal number of clusters US counties
set.seed(1234)
USScaled <- USCovidData %>% column_to_rownames(var = "county_state") %>% scale()
fviz_nbclust(USScaled, kmeans, method = "wss")
```

```{r}
# PCA US counties
res2 <- USCovidData %>% column_to_rownames(var = "county_state") %>% prcomp(scale = TRUE)
get_eig(res2)
df.res2 <- res2$x
```

```{r}
# k-means clustering US counties
set.seed(1234)
kc2 <- df.res2 %>% kmeans(6) 
summary(kc2)
o <- order(kc2$cluster)
data.frame(kc2$cluster[o])
```

```{r warning=FALSE}
# PCA biplot US counties
res2 %>% fviz_pca_biplot(label = "var",
                         habillage = as.factor(kc2$cluster),
                         repel = "TRUE") +
  ylim(-10, 4)
```

```{r}
#Data Visualization
```

```{r}
#Hypothesis Testing
```

```{r}
#Regression
df <- CovidData %>%
  group_by(location, continent) %>%
  summarise(vaccination_received_rate = (max(total_vaccinations, na.rm = TRUE)/max(population, na.rm = TRUE)) * 100,
            total_cases = max(total_cases, na.rm = TRUE),
            total_deaths = max(total_deaths, na.rm = TRUE),
            total_vaccinations = max(total_vaccinations, na.rm = TRUE),
            population = max(population, na.rm = TRUE),
            gdp_per_capita = max(gdp_per_capita, na.rm = TRUE), 
            life_expectancy = max(life_expectancy, na.rm = TRUE), 
            extreme_poverty = max(extreme_poverty, na.rm = TRUE), 
            human_development_index = max(human_development_index, na.rm = TRUE)) %>% 
  na.omit() %>%
  filter(is.finite(vaccination_received_rate),
         is.finite(total_cases),
         is.finite(total_deaths),
         is.finite(total_vaccinations),
         is.finite(population),
         is.finite(gdp_per_capita),
         is.finite(life_expectancy),
         is.finite(extreme_poverty),
         is.finite(human_development_index))
  



head(df)

# data frame with only numericals to perform regression
df.cont <- df %>%
  select(-c(location, continent))

head(df.cont)

```

redo regression
<!-- ```{r} -->
<!-- regfit.full = regsubsets(total_vaccinations ~ ., data = df.cont,  nvmax = 8, method="exhaustive")  -->
<!-- summary(regfit.full) -->
<!-- ``` -->

```{r}

reg.summary <- summary(regfit.full) 

par(mfrow=c(2,2))
#rss plot -  NOT USEFUL
plot(reg.summary$rss ,xlab="Number of Variables ",ylab="RSS",type="l")

#adjr2 plot
plot(reg.summary$adjr2 ,xlab="Number of Variables ", ylab="Adjusted RSq",type="l")
max_adjr2 <- which.max(reg.summary$adjr2)
points(max_adjr2,reg.summary$adjr2[max_adjr2], col="red",cex=2,pch=20)

# AIC criterion (Cp) to minimize
plot(reg.summary$cp ,xlab="Number of Variables ",ylab="Cp", type='l')
min_cp <- which.min(reg.summary$cp )
points(min_cp, reg.summary$cp[min_cp],col="red",cex=2,pch=20)

# BIC criterion to minimize
plot(reg.summary$bic ,xlab="Number of Variables ",ylab="BIC",type='l')
min_bic <- which.min(reg.summary$bic)
points(min_bic,reg.summary$bic[min_bic],col="red",cex=2,pch=20)
```


```{r}
# getting the predictors
x_var <- df.cont %>% select(-vaccination_received_rate) %>% as.matrix()
# getting the independent variable
y_var <- df.cont[,"vaccination_received_rate"]
```

```{r}
# performing lasso regression 

lasso <- glmnet(x_var, y_var, alpha=1)
summary(lasso)
```

```{r}
cv_lasso <- cv.glmnet(x_var, y_var, alpha = 1)
cv_lasso
plot(cv_lasso)
```

```{r}
cv_lasso$lambda.min
cv_lasso$lambda.1se
```

```{r}
lbs_fun <- function(fit, offset_x=1, ...) {
  L <- length(fit$lambda)
  x <- log(fit$lambda[L])+ offset_x
  y <- fit$beta[, L]
  labs <- names(y)
  text(x, y, labels=labs, ...)
}
plot(lasso, xvar = "lambda", label=T)
lbs_fun(lasso)
abline(v=log(cv_lasso$lambda.min), col = "red", lty=2)
abline(v=log(cv_lasso$lambda.1se), col="blue", lty=2)
```

```{r}
min_lasso <- glmnet(x_var, y_var, alpha=1, lambda= cv_lasso$lambda.min)
coef(min_lasso)

se_lasso <- glmnet(x_var, y_var, alpha=1, lambda= cv_lasso$lambda.1se)
coef(se_lasso)
```

```{r}
# Make predictions on the test data
predictions <- min_lasso %>% predict(x_var) %>% as.vector()
# Model performance metrics
data.frame(
  RMSE = RMSE(predictions, y_var),
  Rsquare = R2(predictions, y_var)
)
```

```{r}
## note: 

# LASSO suggests model with 2 predictors

# AIC and BIC suugests model with 3 predictors

# from LASSO, it suggests that the best model is one with predictor gdp_per_capita and life_expectancy.
# which both predictors are highly correlated? 
```

```{r}
# simple linear regression

# using df.cont

attach(df.cont)
lm.fit1 <- lm(vaccination_received_rate ~ ., data = df.cont)
summary(lm.fit1) 

# 3 significant variables: gdp_per_capita, human_development_index, life_expectancy; as suggested by BIC and AIC
```

```{r}
library(car)
vif(lm.fit1)

plot(lm.fit1) # passed all assumptions
```

```{r}
# a good vif values are those within 10

# those are: gdp_per_capita, extreme_poverty, life expectancy

lm.fit2 <- lm(vaccination_received_rate ~ gdp_per_capita + extreme_poverty + life_expectancy, data = df.cont)
summary(lm.fit2) 
```

```{r}
# simple linear regresion 

lm.fit3 <- lm(vaccination_received_rate ~life_expectancy, data = df.cont)
summary(lm.fit3) 
```

```{r}
lm.fit3 <- lm(vaccination_received_rate ~gdp_per_capita, data = df.cont)
summary(lm.fit3) 
```

```{r}
# performing regression using the model suggested by the lasso

lm.fit4 <- lm(vaccination_received_rate ~gdp_per_capita + life_expectancy, data = df.cont)
summary(lm.fit4) 
```

```{r}
AIC(lm.fit4)
```


```{r}
# intuition: country with high life expentancy is modern countries.

# so let's check its correlation
cor(life_expectancy, gdp_per_capita)
```

### Conclusion

- We have statistical evidence that countries with high life expentancy has receives more covid vaccines. 

- Make sense as most of these countries produce the vaccines, and have more money to buy vaccines. 

- Those country with high life expectancy are first class world countries. 


```
